Hijacking Simulators with Universal Probabilistic Programming

### 2019-07-11 paper/learning/notes.tex
updated

One challenge in performing inference in such models is the combinatorial increase in \emph{paths}
that can be taken within the given probabilistic model.
This is because each member of the population has 
a bounded, but large number of decisions to make. 
In addition to this, each member of the population can interact
with other members of the population.
To model a single population member quantum-based methods leveraging 
linear superposition and entanglement provide an ability to explore multiple paths simultaneously maintaining 
the whole program, \emph{state}, in one iteration.
 Unfortunately, analytically writing the functional 
form of the state is, in this instance, impossible.   
The usability of current classical inference methods in this context is also problematic. 
However, by combining hijacking and amortizing parts of the simulator, in particular rejection sampling 
loops, we can not only correctly perform inference in such models, as the rejection sampling loops are 
now replaced with learnt functions, but we can also perform inference much more efficiently. 
This is because the rejection sampling loops are now replaced with learnt functions which enables us to correctly
construct the density and as they are amortized evaluation is cheap and only has to be performed once per call.
This both reduces memory consumption and running time.

As updates are made sequentially MCMC methods typically are not-well suited this type of recursive
estimation problem. 
As each time we get a new data point $y_{t+1}$ we have to run new MCMC simulations 
for $P(x_{0:t_1}| y_{0:t+1})$, we also cannot directly reuse the samples $\{x^{i}_{0:t}\}$. 
Likewise, importance sampling is not well suited either, as we cannot reuse the samples and weights for a time $t$ $\{x^{i}_{0:t}, \tilde{w}^{i}_{t}\}^{N}_{i=1}$ 
to sample from $(x_{0:t_1}| y_{0:t+1})$.
 However, as we can only run the simulator $\texttt{forward()}$ we cannot return back to a previous state and re-run. So adapting existing inference techniques 
 is the only viable option.

@@

When the number of observations is fixed, the combined weight is equal to $w(x_{1:n_x}) = \prod^{n_y}_{j=1} w_j $. 
To generate the weight up to each observe we can define $K(j)$ to be a function that returns the least $i$ before hitting an observe $j$.
This enables us to write the $j$-th weight as follows:
\begin{equation}
  w_j(x1:K(j)) = g_{b_j}(y_j | \psi_j) \prod^{K(j)}_{K(j-1)+1} v_i
\end{equation}
where we define $v_j = \frac{f_{a_j}(x_j | \eta_j)}{q_j(x_j |f_{a_j}, \eta_j)}$ \textbf{Note: Ask Tom why the proposal is also dependent
on $f_{a_i}$. }

Taking these definitions we can construct a target density over the 
latents that is proportional to the products of \texttt{sample} and \texttt{observe}
statements \textbf{Note: Add update posterior here}
Thus, given the proposal $q(x_{1:K(t)})$ we define the importance weight as:
\begin{multline}
  w(x_{1:t}) = \prod^{t}_{j=1} w_j (x_{1:K(j)}) = \\
   \frac{p_t(x_{1:K(t)})}{q(x_{1:K(t)})} =
 \begin{cases}
  \prod^{t}_{j=1}g_{b_j}(y_j | w_i)\prod^K(j)_{K(j-1)+1} v_i \hspace{0.1cm} \texttt{if} \hspace{0.1cm} B_t(x_1:K(t))=1 \\
  0
 \end{cases}
\end{multline}

@@

@@

Due to the current set-up performing Random-walk Metropolis Hastings (RMH) is not feasible 
as the rejection sampling blocks within the simulator mean that we cannot correctly perform RMH
as we do not have a functional form for all parts of the simulator \textbf{Note: Re Word}

\section{An Example: Modelling Parasite Densities in Simulation}

\subsection{The Model}

The parasite model by Smith el al.~\cite{smith2006relationship} is a simple stochastic 
model for determining the number of infections within a given population demographic.
The models hypothesised in ~\cite{smith2006relationship} were found, 
after extensive testing and development, to not characterise well the number 
of infections among different demographics, especially those under 15 years old. 
By using probabilistic programming we can immediately understand 
the expressiveness of the proposed model and can exact important insights for 
extending the proposed models, which increases ones ability to develop 
improved models to estimate parasite densities, given 
our observations of the seasonal entomological inoculation rate (EIR).
In addition to this we can greatly reduce the computational costs, whilst
improving model comprehension and expressivity.  
Although a statistical model is not formally defined within~\cite{smith2006relationship}
we define the unknown \emph{disease dynamics} parameters with a set of latent variables$\{x_{i}\}^{N}$
and construct the model as follows:

\begin{align}
  \mu_i &= N \\ % pop size
  % disease dynamics
  S_{\infty} &\sim \mathcal{N}(\mu_\infty, \sigma^{2}_\infty) \\
  E_{crit} &\sim \mathcal{N}(\mu_{crit}, \sigma^{2}_{crit}) \\
  S_{imm} &\sim \mathcal{N}(\mu_{imm}, \sigma^{2}_{imm}) \\
  X_{p_{crit}} &\sim \mathcal{N}(\mu_{p_{crit}}, \sigma^{2}_{p_{crit}}) \\
  \gamma_p &\sim \mathcal{N}(\mu_{p}, \sigma^{2}_{p}) \\
  \rho_i & \sim \mathcal{U}(\mathbb{L}_i , \mathbb{U}_i) \\
  % integration for parasite densities in different demographics
  for \hspace{0.1cm} t &= 1:t_{max}: \\
  E_a &= \frac{E_{{max}_t}}{A_{max}} \times h_{\mu_{SA}}() \\
  X_p &= E_a + X_p \\
  S_1 &= S_{\infty} + \frac{(1 - S_\infty) }{1 + \frac{E_a}{E_{crit}}} \\
  S_2 &= S_{imm} + \frac{1 - S_{imm}}{1 + (\frac{X_p}{X_p_{crit}})^{\gamma_p}}\\
  %Probability of survival function
  S_p &= S_1 \times S_2 \\
  \lambda &= S_p \times E_a \\
  % Number of infections generated in unit time per individual sampled from poisson
  h &\sim Pos(\lambda) \\
  \texttt{if} \hspace{0.1cm} h > 0: \\
  \hspace{2.3cm} n_{infect} = 1
  % % end for - add prevelance construction below
  % \mathcal{P}rev &= 
\end{align}

\begin{figure}[ht!]
  \centering
  \label{fig:parasite}
  \includegraphics[width=0.5\textwidth]{parasite_model.pdf}
  \caption{The DAG of the Parasite model.}
\end{figure}

\begin{figure}[hb!]
  \label{fig:amortized_sampling}
  \includegraphics[width=0.5\textwidth,height=0.5\textheight]{amoritized_rejection.pdf}
  \caption{Using PyProb we can locate all regions of inefficient sampling 
  and replace those points in the code with learnt functions that represent
  the distributions of the rejection sampling statements. }
\end{figure}

\bibliographystyle{plain}
\bibliography{refs}

>>>>

One challenge in performing inference in such models is the combinatorial increase in \emph{paths}
that can be taken within the given probabilistic model.
This is because each member of the population has 
a bounded, but large number of decisions to make. 
In addition to this, each member of the population can interact
with other members of the population.
To model a single population member quantum-based methods leveraging 
linear superposition and entanglement provide an ability to explore multiple paths simultaneously maintaining 
the whole program, \emph{state}, in one iteration.
 Unfortunately, analytically writing the functional 
form of the state is, in this instance, impossible.   
The usability of current classical inference methods in this context is also problematic. 
However, by combining hijacking and amortizing parts of the simulator, in particular rejection sampling 
loops, we can not only correctly perform inference in such models, as the rejection sampling loops are 
now replaced with learnt functions, but we can also perform inference much more efficiently. 
This is because the rejection sampling loops are now replaced with learnt functions which enables us to correctly
construct the density and as they are amortized, evaluation is cheap and only has to be performed once per call.
This significantly reduces memory consumption and running time.

Without the amortization procedure, as updates are made sequentially MCMC methods are typically not-well suited this type of recursive
estimation problem. 
As each time we get a new data point $y_{t+1}$ we have to run new MCMC simulations 
for $P(x_{0:t_1}| y_{0:t+1})$, we also cannot directly reuse the samples $\{x^{i}_{0:t}\}$. 
However, as we can only run the simulator $\texttt{forward()}$ we cannot return back to a previous state and re-run. 
So adapting existing inference techniques is the only viable option. 
MCMC methods can be effective with a well designed amortized inference procedure (\textbf{Note: This is not guaranteed}).
Non-MCMC based methods such as importance sampling (IS) and sequential monte carlo (SMC), may also be
sensible inference engines, although we shall discuss the limitations within the simulator context in the proceeding sections.

However, to implement this in practice we are required to do the following:
 \begin{itemize}
    \item Create a sensible training procedure
    \item Have 2 additional functions implemented within pyprob
    \begin{enumerate} \item 1) That stores only the necessary raw samples from the desired rejection sampling addresses to be used for training.
                       \item 2) A function at \emph{test time} that replaces those given addresses with a learnt surrogate.
    \end{enumerate}
    \item Finer grained control over the actual latent variables of interest, to avoid unnecessary memory consumption and ensure that inference problems remain tractable.
  \end{itemize}
There is no doubt that implementing some fo these features will be more invasive to the underlying simulator code than previously
just hijacking the raw-random number draws, but, it would enable us to attack difficulties in the inference procedures that are
currently not amenable to the current approach.

@@

When the number of observations is fixed, the combined weight is equal to $w(x_{1:n_x}) = \prod^{n_y}_{j=1} w_j $. 
To generate the weight up to each observe we can define $K(j)$ to be a function that returns the least $i$ before hitting an observe $j$.
This enables us to write the $j$-th weight as follows:
\begin{equation}
  w_j(x1:K(j)) = g_{b_j}(y_j | \psi_j) \prod^{K(j)}_{K(j-1)+1} v_i
\end{equation}
where we define $v_j = \frac{f_{a_j}(x_j | \eta_j)}{q_j(x_j |f_{a_j}, \eta_j)}$

Taking these definitions we can construct a target density over the 
latents that is proportional to the products of \texttt{sample} and \texttt{observe}
statements \textbf{Note: Add update posterior here}.
Thus, given the proposal $q(x_{1:K(t)})$ we define the importance weight as:
\begin{multline}
  w(x_{1:t}) = \prod^{t}_{j=1} w_j (x_{1:K(j)}) = \\
   \frac{p_t(x_{1:K(t)})}{q(x_{1:K(t)})} =
 \begin{cases}
  \prod^{t}_{j=1}g_{b_j}(y_j | w_i)\prod^K(j)_{K(j-1)+1} v_i \hspace{0.1cm} \texttt{if} \hspace{0.1cm} B_t(x_1:K(t))=1 \\
  0
 \end{cases}
\end{multline}

One problem with importance sampling is in its ability to scale and condition upon multiple observations. |
Choosing a \emph{good} prior remains challenging.

@@

To implement SMC in the current set-up we round require a forking method and the

@@

Due to the current set-up performing Random-walk Metropolis Hastings (RMH) is not feasible 
as the rejection sampling blocks within the simulator mean that we cannot correctly perform RMH
as we do not have a functional form for all parts of the simulator to construct a target density
that we can evaluate. \textbf{Note:
Ask tom what:  one can sometimes achieve improvements by instead using the type of the distribution object 
associated with $x_m$ (what does this first part mean?) 
 to automatically construct a valid random walk proposal that allows for improved
 hill climbing behavior on the individual updates. }.

\section{An Example: Modelling Parasite Densities in Simulation}
% \begin{wrapfigure}{r}{0.4\textwidth}
  \begin{figure}[h!]
    \centering
    \label{fig:parasite}
    \includegraphics[width=0.4\textwidth]{parasite_model.pdf}
    \caption{The DAG of the Parasite model.}
  % \end{wrapfigure}
  \end{figure}
The parasite model by Smith el al.~\cite{smith2006relationship} is a simple stochastic 
model for determining the number of infections within a given population demographic.
The models hypothesised in ~\cite{smith2006relationship} were found, 
after extensive testing and development, to not characterise well the number 
of infections among different demographics, especially those under 15 years old. 
By using probabilistic programming we can immediately understand 
the expressiveness of the proposed model and can exact important insights for 
extending the proposed models, which increases ones ability to develop 
improved models to estimate parasite densities, given 
our observations of the seasonal entomological inoculation rate (EIR).
In addition to this we can greatly reduce the computational costs, whilst
improving model comprehension and expressivity.  
Although a statistical model is not formally defined within~\cite{smith2006relationship}
we define the unknown \emph{disease dynamics} parameters with a set of latent variables$\{x_{i}\}^{N}$
and construct the model as follows:

\begin{align}
  \mu_i &= N \\ % pop size
  % disease dynamics
  S_{\infty} &\sim \mathcal{N}(\mu_\infty, \sigma^{2}_\infty) \\
  E_{crit} &\sim \mathcal{N}(\mu_{crit}, \sigma^{2}_{crit}) \\
  S_{imm} &\sim \mathcal{N}(\mu_{imm}, \sigma^{2}_{imm}) \\
  X_{p_{crit}} &\sim \mathcal{N}(\mu_{p_{crit}}, \sigma^{2}_{p_{crit}}) \\
  \gamma_p &\sim \mathcal{N}(\mu_{p}, \sigma^{2}_{p}) \\
  \rho_i & \sim \mathcal{U}(\mathbb{L}_i , \mathbb{U}_i) \\
  % integration for parasite densities in different demographics
  for \hspace{0.1cm} t &= 1:t_{max}: \\
  E_a &= \frac{E_{{max}_t}}{A_{max}} \times h_{\mu_{SA}}() \\
  X_p &= E_a + X_p \\
  S_1 &= S_{\infty} + \frac{(1 - S_\infty) }{1 + \frac{E_a}{E_{crit}}} \\
  S_2 &= S_{imm} + \frac{1 - S_{imm}}{1 + (\frac{X_p}{X_p_{crit}})^{\gamma_p}}\\
  %Probability of survival function
  S_p &= S_1 \times S_2 \\
  \lambda &= S_p \times E_a \\
  % Number of infections generated in unit time per individual sampled from poisson
  h &\sim Pos(\lambda) \\
  \texttt{if} &\hspace{0.1cm} h > 0: \\
  \hspace{2.3cm} &n_{infect} = 1 \\
  \text{TODO: add prevelance construction below}
  % % end for - add prevelance construction below
  % \mathcal{P}rev &= 
\end{align}

\textbf{TODO: Add some of the plots generated from this model and an accompanying analysis to them.}

\begin{wrapfigure}{l}{0.5\textwidth}
  \label{fig:amortized_sampling}
  \includegraphics[width=0.5\textwidth,height=0.5\textheight]{amoritized_rejection.pdf}
  \vspace{-50pt}
  \caption{Using PyProb we can locate all regions of inefficient sampling 
  and replace those points in the code with learnt functions that represent
  the distributions of the rejection sampling statements. $Z_{i}$ represents the marginal and 
  by learning surrogate functions for each of the rejection sampling loops.
  In doing so we can learn approximate raw sample distributions $f(x_{i} | \eta)$ for each
  rejection sampling loop. This means that we can evaluate the function at each part in the simulator, enabling
  us to construct a target density that can be evaluated when using MCMC inference schemes such as RMH.
    }
  \end{wrapfigure}

\section{Additional TODOs}

\begin{itemize}
\item Before implementing such tools into the simulator, spend time looking for simpler stochastic simulators 
that can be hijacked but have target densities that can be constructed analytically. 
\item If we are not utilizing the additional information stored in the trace, which is only used during the inference compilation
training procedure, then we do not need it all, as for importance sampling we only require the weights after each \texttt{forward()}
update. 
\item If we cannot find any pre-existing simulators that are amenable to this approach then, we need to create our own, with
a rejection sampling loop, that can be exploited by the new machinery. This would enable us to debug, provide a simple example 
for the paper and understand the feasibility of the entire approach. 
\end{itemize}
\bibliographystyle{plain}
\bibliography{refs}

### 2019-07-11 paper/learning/thesystem.tex
updated

\begin{tikzpicture}[->,>=stealth']

 % Position of Sample 
 % Use previously defined 'state' as layout (see above)
 % use tabular for content to get columns/rows
 % parbox to limit width of the listing
 \node[state,
 text width=3cm, 	% max text width
 yshift=0cm, 		% move 2cm in y
 node distance=0cm, 	% distance to Sample
 anchor=center] (Sample) 
 {\begin{tabular}{l}{
  \textbf{Sample}\\
  \parbox{4cm}{
    \small{Program call:\\
    \emph{Program}.\texttt{forward()}}
  }
 \end{tabular}};

@@

 % STATE EPC
 \node[state,
  right of=System,
  yshift = 1cm,
  node distance=5cm,
  anchor=center] (EPC) 
 {%
 \begin{tabular}{l}
  \textbf{Program density}\\
  \parbox{5cm}{P(\mathcal{T}= $x_{1:n_x}| \lambda)$\\
  \{$x_{1:n_{x}}\}$,$w= \prod^{n_{y}}_{i=1}g_b_i (y_i | \psi_i)}$}
 \end{tabular}
 };

@@

\end{tikzpicture}

>>>>

\begin{figure}
\begin{tikzpicture}[->,>=stealth']

 % Position of Sample 
 % Use previously defined 'state' as layout (see above)
 % use tabular for content to get columns/rows
 % parbox to limit width of the listing
 \node[state,
 text width=3cm, 	% max text width
 yshift=0cm, 		% move 2cm in y
 node distance=0cm, 	% distance to Sample
 anchor=center] (Sample) 
{%
\begin{tabular}{l} 	% content
  \textbf{Sample}\\
  \parbox{4cm}{
   \small{Program call:\\
   \emph{Program}.\texttt{forward()}}
 }
 \end{tabular}};

@@

 % STATE EPC
 \node[state,
  right of=System,
  yshift = 1cm,
  node distance=5cm,
  anchor=center] (EPC) 
 {%
 \begin{tabular}{l}
  \textbf{Program density}\\
  \parbox{5cm}{P($\mathcal{T}= x_{1:n_x}| \lambda)$\\
  \{$x_{1:n_{x}}\}$,$w= \prod^{n_{y}}_{i=1} g_{b_i}(y_i | \psi_i)$}
 \end{tabular}
 };

@@

\end{tikzpicture}
\caption{An overview of the hijacking system. $x_i$ represent latent variables. $f_{a_i}$ represent raw sample calls,
$g_{b_j}$ represent conditioning statements. $\mathcal{T}$ corresponds to the program trace. $w$ represents the collective weights. 
$r$ represents run number. $n_x$ represents the number of latent variables and $n_y$ represents the number of
observations, which will always be fixed, but is model dependent. $\lambda$ represents all the other variables generated 
by running the simulator \texttt{.forward()}.
}
\label{fig:systemoverview}
\end{figure}
